---
title: "Merging Data Discovery Day"
author: "Halle Prine"
format: html
editor: visual
---

This document is the r-code I will be using to clean and merge the data I need for Discovery Day 2025!

*Loading in the Data*
```{r include=FALSE}
# The "here" library will let me read the data from an individual folder.The dplyr will allow me to clean the data by sorting out participants that have less than 17 days of daily or bed data. 

install.packages("dplyr")
install.packages("here")

library(dplyr)
library(here)

# I read the data and put it in a data frame object.

baseline <- read.csv(here("Data", "Baselinesurvey.CSV"), header = FALSE, sep = ",") 

daily <- read.csv(here("Data", "DailySummaries_2025.CSV"), header = FALSE, sep = ",") 

bed <- read.csv(here("Data", "DailyTimeInBed_2025.CSV"), header = FALSE, sep = ",") 

moca <- read.csv(here("Data", "MOCA Scores.CSV"), header = FALSE, sep = ",")


```


Starting with Baseline Data I will rename and clean the variables. 

```{r include=FALSE}
# Remove the first, second, and third rows but keep everything below
baseline <- baseline[-c(1,2,3), ]

# Reset row indices
rownames(baseline) <- NULL

View(baseline)

# Here we are re-naming our essential variables we want from the Baseline data
names(baseline)[names(baseline)=="V18"]<-"ID"   
names(baseline)[names(baseline)=="V19"]<-"DOB"   
names(baseline)[names(baseline)=="V20"]<-"sex"   
names(baseline)[names(baseline)=="V21"]<-"ethnicity"   
names(baseline)[names(baseline)=="V22"]<-"race"   
names(baseline)[names(baseline)=="V23"]<-"marital.st"   
names(baseline)[names(baseline)=="V24"]<-"alone" 
names(baseline)[names(baseline)=="V25"]<-"occupation"   
names(baseline)[names(baseline)=="V26"]<-"education" 
names(baseline)[names(baseline)=="V27"]<-"research.exp"   
names(baseline)[names(baseline)=="V28"]<-"research.time"  
```


I will do the same thing to the Pal-daily summaries and time in bed data. 

```{r include=FALSE}

# Here we are re-naming our variables we want from the Pal-daily summaries data
names(daily)[names(daily)=="V1"]<-"Folder"
names(daily)[names(daily)=="V2"]<-"Filename"
names(daily)[names(daily)=="V3"]<-"Serial Number"
names(daily)[names(daily)=="V4"]<-"File Code"
names(daily)[names(daily)=="V5"]<-"ID" 
names(daily)[names(daily)=="V6"]<-"Day of Week"
names(daily)[names(daily)=="V7"]<-"Date"
names(daily)[names(daily)=="V8"]<-"Valid Day"
names(daily)[names(daily)=="V9"]<-"Inverted wear correction"
names(daily)[names(daily)=="V10"]<-"Alignment (%)"
names(daily)[names(daily)=="V11"]<-"Total Time (m)"
names(daily)[names(daily)=="V12"]<-"Data Errors (m)"
names(daily)[names(daily)=="V13"]<-"Total Num Steps"
names(daily)[names(daily)=="V14"]<-"Num Cycling Steps"
names(daily)[names(daily)=="V15"]<-"Upright Time(m)"
names(daily)[names(daily)=="V16"]<-"Standing Time(m)"
names(daily)[names(daily)=="V17"]<-"Total Stepping Time(m)"
names(daily)[names(daily)=="V18"]<-"Cycling Time(m)"
names(daily)[names(daily)=="V19"]<-"Sitting Time(m)"
names(daily)[names(daily)=="V20"]<-"Seated Transport Time(m)"
names(daily)[names(daily)=="V21"]<-"Primary Lying Time(m)"
names(daily)[names(daily)=="V22"]<-"Secondary Lying Time(m)"
names(daily)[names(daily)=="V23"]<-"Non Wear Time(m)"
names(daily)[names(daily)=="V24"]<-"Num Sit To Stand Transitions"
names(daily)[names(daily)=="V25"]<-"Num Stand To Sit Transitions"
names(daily)[names(daily)=="V26"]<-"MET"
names(daily)[names(daily)=="V27"]<-"Num Sitting Bouts >30m"
names(daily)[names(daily)=="V28"]<-"Num Sitting Bouts >60m"
names(daily)[names(daily)=="V29"]<-"Time Spent In Sitting Bouts >30m"
names(daily)[names(daily)=="V30"]<-"Time Spent In Sitting Bouts >60m"
names(daily)[names(daily)=="V31"]<-"Stepping Time(m) (duration <1m)"
names(daily)[names(daily)=="V32"]<-"Stepping Time(m) (duration >1m <=5m)"
names(daily)[names(daily)=="V33"]<-"Stepping Time(m) (duration >5m <=10m)"
names(daily)[names(daily)=="V34"]<-"Stepping Time(m) (duration >10m <=20m)"
names(daily)[names(daily)=="V35"]<-"Stepping Time(m) (duration >20m)"
names(daily)[names(daily)=="V36"]<-"Num Steps (duration <=1m)"
names(daily)[names(daily)=="V37"]<-"Num Steps (duration >1m <=5m)"
names(daily)[names(daily)=="V38"]<-"Num Steps (duration >5m <=10m)"
names(daily)[names(daily)=="V39"]<-"Num Steps (duration >10m <=20m)"
names(daily)[names(daily)=="V40"]<-"Num Steps (duration >20m)"
names(daily)[names(daily)=="V41"]<-"Stepping Time(m) (Cadence>=75)"
names(daily)[names(daily)=="V42"]<-"Stepping Time(m) (Cadence>=75, duration>1m)"
names(daily)[names(daily)=="V43"]<-"Num Steps (Cadence>=75)"
names(daily)[names(daily)=="V44"]<-"Num Steps (Cadence>=75, duration>1m)"
names(daily)[names(daily)=="V45"]<-"ST_100Cad"
names(daily)[names(daily)=="V46"]<-"ST_100Cad_1m"
names(daily)[names(daily)=="V47"]<-"NS_100Cad"
names(daily)[names(daily)=="V48"]<-"NS_100Cad_1m"

daily <- daily[-1, ]  
rownames(daily) <- NULL  # Reset row indices
colnames(daily)


View(daily)

# Here we are re-naming our variables we want from the Time in Bed data
names(bed)[names(bed)=="V1"]<-"Folder"
names(bed)[names(bed)=="V2"]<-"Filename"
names(bed)[names(bed)=="V3"]<-"Serial Number"
names(bed)[names(bed)=="V4"]<-"File Code"
names(bed)[names(bed)=="V5"]<-"ID" 
names(bed)[names(bed)=="V6"]<-"Waking Day Before TIB"
names(bed)[names(bed)=="V7"]<-"Waking Day After TIB"
names(bed)[names(bed)=="V8"]<-"Week Days"
names(bed)[names(bed)=="V9"]<-"Valid Day"
names(bed)[names(bed)=="V10"]<-"TIB Start Date"
names(bed)[names(bed)=="V11"]<-"TIB Start Time"
names(bed)[names(bed)=="V12"]<-"Date"
names(bed)[names(bed)=="V13"]<-"TIB End Time"
names(bed)[names(bed)=="V14"]<-"TIB duration (h)"

bed <- bed[-1, ]  
rownames(bed) <- NULL  # Reset row indices

View(bed)
```

We wont need to re-name any variables within the MOCA data file since I've 
cleaned the data by hand into an excel sheet from SPARC's measurement Tab. 


Now I will go into the Daily and Bed data from the ActivPal and take out any cases 
that have less than 17 days of ActivPal data. I will also cut data that has more 
than 20 days to exactly 20 days. 

```{r include=FALSE}

# The code below wasn't running so I checked the data frame's columns names to 
# check is "ID" is in the data frame.
colnames(daily) # It is in the data frame

#Now I'm going to check if there are values in the ID column and check the class
head(daily$ID) # There are values
class(daily$ID) # I see that it's classified as a character so I need to make it 
# a factor. 

daily$ID <- as.factor(daily$ID) # Making ID a factor
bed$ID <- as.factor(bed$ID)
# Count the number of days each participant has data for in 'daily' and 'bed'

daily_filtered <- daily %>%
  group_by(ID) %>%
  filter(n() >= 17) %>%  # Keep IDs with at least 17 days
  arrange(Date) %>%  # Arrange by Date (optional but helpful)
  filter(row_number() <= 20) %>%  # Keep only the first 20 rows per ID
  ungroup()

bed_filtered <- bed %>%
  group_by(ID) %>%
  filter(n() >= 17) %>%  # Keep IDs with at least 17 days
  arrange(Date) %>%  # Arrange by Date (optional)
  filter(row_number() <= 20) %>%  # Keep only the first 20 rows per ID
  ungroup()


# remove any duplicates 
daily_filtered <- daily_filtered %>%
  group_by(ID, Date) %>%
  slice_head(n = 1) %>%  # Keep the first row per ID-Date
  ungroup()

bed_filtered <- bed_filtered %>%
  group_by(ID, Date) %>%
  slice_head(n = 1) %>%  # Keep the first row per ID-Date
  ungroup()


```


Time to merge ActivPal Data together from the daily summaries and time in bed data

```{r include=FALSE}

# We will be left joining the two by ID and Date just double checking if date was
# the same class between the two data frames. 

class(bed_filtered$Date)

class(daily_filtered$Date)

pal <- left_join(bed_filtered, daily_filtered, by = c("ID", "Date"))
View(pal)

pal %>%
  count(ID) %>%
  print()

```


Now I need to clean and rename columns in the moca data frame

```{r include=FALSE}

View(moca)

# rename the columns 
names(moca)[names(moca)=="V1"]<-"ID"
names(moca)[names(moca)=="V2"]<-"Date.MoCA_1"
names(moca)[names(moca)=="V3"]<-"MoCA_1"
names(moca)[names(moca)=="V4"]<-"Date.MoCA_2"
names(moca)[names(moca)=="V5"]<-"MoCA_2"

moca <- moca[, -6]
moca <- moca[-1, ] 
```


Now I want to filter the MoCA ID's to match the Pal data ID's so we can merge 
the two data frames. 

```{r include=FALSE}

# Filter moca to include only IDs that are in pal
moca_filtered <- moca %>%
  filter(ID %in% unique(pal$ID))

View(moca_filtered)

# Join the repeated MoCA data with pal
pal_moca <- left_join(pal, moca, by = "ID")

# View the result
pal_moca %>%
  group_by(ID) %>%
  tally()  # Should return ≈20 per ID

View(pal_moca)

write.csv(pal_moca, "pal_moca.csv", row.names = FALSE)

```


Ok...now for the final merge with the baseline data and the pal_moca data frame.

```{r include=FALSE}
# Check class on baseline data frame's ID and make it a factor so we can merge
class(baseline$ID)
baseline$ID <- as.factor(baseline$ID)

# Cut the ID's from baseline that are not included in the final analyses
baseline_filtered <- baseline %>%
  filter(ID %in% pal_moca$ID)

baseline_unique <- baseline_filtered %>%
  group_by(ID) %>%
  slice(1) %>%
  ungroup()

# Now do the final merge
discover25.dta <- left_join(pal_moca, baseline_unique, by = "ID")

# check for unintentional duplication
table(duplicated(discover25.dta$ID))  # Should return FALSE for all

count(discover25.dta, ID)  # Should still show 20 rows per ID
View(discover25.dta)

write.csv(discover25.dta, "discover25_dta.csv", row.names = FALSE)

```
